Efficient implementation of penalized regression\\for genetic risk prediction

### 2018-09-21 paper/draft.tex

\abstract{
Polygenic Risk Scores (PRS) consist in combining the information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease. PRS might have a major public health impact, possibly allowing for screening campaigns to identify high-genetic risk individuals for a given disease.  
The ``Clumping+Thresholding'' (C+T) approach, which is the most common method to derive PRS, uses only univariate genome-wide association studies (GWAS) summary statistics, which makes it fast and easy to use. 
However, previous work showed that jointly estimating SNP effects for computing PRS has the potential to significantly improve the predictive performance of PRS as compared to C+T.

@@

In conclusion, our study demonstrates that penalized logistic regression is applicable to large-scale individual-level data and can achieve more discriminative polygenic risk scores. Our implementation is publicly available in R package bigstatsr.\\

\textbf{Contact:} \href{florian.prive@univ-grenoble-alpes.fr}{florian.prive@univ-grenoble-alpes.fr} \& \href{michael.blum@univ-grenoble-alpes.fr}{michael.blum@univ-grenoble-alpes.fr}\\
\textbf{Supplementary information:}
}

@@

\newpage
\section{Introduction}

@@

Several methods have been developed to predict disease status, or more generally any phenotype, based on SNP information. A commonly used method often called ``P+T'' or ``C+T'' (which stands for ``Clumping and Thresholding'') is used to derive PRS from results of Genome-Wide Association Studies (GWAS) \cite[]{chatterjee2013projecting,dudbridge2013power,evans2009harnessing,purcell2009common,wray2007prediction}. This technique uses GWAS summary statistics only, allowing for fast implementation. However, the ``C+T'' approach also has several limitations. 
Previous studies have shown that predictive performance of C+T is very sensitive to the threshold of inclusion of SNPs, depending on the disease architecture \cite[]{ware2017heterogeneity}.
Linear Mixed-Models (LMMs) are another widely-used method in fields such as plant and animal breeding or for predicting highly heritable quantitative human phenotypes such as height \cite[]{lello2017accurate,yang2010common}. Yet, models resulting from LMM, known e.g.\ as ``gBLUP'', are not optimal for predicting disease status based on genotypes \cite[]{abraham2013performance}. Moreover, these methods and their derivatives are often computationally demanding, both in terms of memory and time required, which makes them unlikely to be used for prediction on very large datasets \cite[]{golan2014effective,maier2015joint,speed2014multiblup,zhou2013polygenic}.
Finally, statistical learning methods have also been used to derive PRS for complex human diseases by jointly estimating SNP effects. Such methods include joint logistic regression, Support Vector Machine (SVM) and random forests \cite[]{abraham2012sparsnp,abraham2014accurate,botta2014exploiting,okser2014regularized,wei2009disease}.

@@

To perform simulations, we use real genotype data and simulate new phenotypes. 
In order to make our comparison as comprehensive as possible, we compare different disease architectures by varying the number, size and location of causal effects as well as the disease heritability. We also compare different models for simulating phenotypes, one with only linear effects, and one that combines linear, dominant and interaction-type effects.
Overall, we find that the penalized logistic regression consistently achieves higher predictive performance than the C+T and T-Trees methods while being as fast as C+T. This demonstrates the feasibility and relevance of this approach for PRS computation on large modern datasets.

@@

We use real genotypes of European individuals from a case-control celiac disease study \cite[]{dubois2010multiple}. The composition of this dataset is presented in table \ref{tab:celiac-data}. Details of quality control and imputation for this dataset are available in \cite{prive2017efficient}. For simulations presented later, we first restrict this dataset to controls from UK in order to remove the genetic structure induced by the celiac disease status and population structure. This filtering process results in a sample of 7100 individuals (see supplementary notebook ``preprocessing''). 
We also use this dataset for real data application, in this case keeping all 15,155 individuals (4496 cases and 10,659 controls). Both datasets contain 281,122 SNPs.

@@

We simulate binary phenotypes using a Liability Threshold Model (LTM) with a prevalence of 30\% \cite[]{falconer1965inheritance}. We vary parameters of the simulations in order to match a range of genetic architecture from low to high polygenicity. 
This is achieved by varying the number of causal variants and their location (30, 300, or 3000 anywhere in all 22 chromosomes or 30 in the HLA region of chromosome 6), and the disease heritability (50\% or 80\%). 
%SNP effects are drawn from either a Gaussian or from a Laplace distribution.
Liability scores are computed either from an additive model with linear effects only (``ADD'') or a more complex model that combines linear, dominant and interaction-type effects (``COMP''). For model ``ADD'', we compute the liability score of the i-th individual $$y_i = \sum_{j\in S_\text{causal}} w_j \cdot \widetilde{G_{i,j}} ~~+~~ \epsilon_i~,$$ where $w_j$ are weights generated from a Gaussian or a Laplace distribution, $G_{i,j}$ is the allele count of individual $i$ for SNP $j$, $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs), $\epsilon$ follows a Gaussian distribution $N(0, 1 - h^2)$ and $S_\text{causal}$ is the set of causal SNPs.
For model ``COMP'', we simulate liability scores using linear, dominant and interaction-type effects (see Supplementary Materials).

@@

In this study, we compare three different types of methods: the C+T method, T-Trees and penalized logistic regression.

The C+T (Clumping + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of Genome-Wide Associations Studies (GWAS). In GWAS, a coefficient of regression (i.e.\ the estimated effect size $\hat\beta_j$) is learned independently for each SNP $j$ along with a corresponding p-value $p_j$. The SNPs are first clumped (C) so that there remain only loci that are weakly correlated with one another (this set of SNPs is denoted $S_\text{clumping}$). Then, thresholding (T) consists in removing SNPs with p-values larger than a threshold $p_T$ to be determined. Finally, a PRS is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients $$\rm{PRS}_i = \sum_{\substack{j \in S_\text{clumping} \\ p_j~<~p_T}} \hat\beta_j \cdot G_{i,j}~,$$ where $\hat\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS. In this study, we mostly report scores for a clumping threshold at $r^2 > 0.2$ within regions of 500kb, but we also investigate thresholds of $0.05$ and $0.8$. We report three different scores of prediction: one including all the SNPs remaining after clumping (denoted ``PRS-all''), one including only SNPs remaining after clumping and that have a p-value under the GWAS threshold of significance ($p < 5 \cdot 10^{-8}$, ``PRS-stringent''), and one that maximizes the AUC (``PRS-max'') for these two thresholds ($0$ and $5 \cdot 10^{-8}$) and a sequence of 100 values of thresholds ranging from $10^{-0.1}$ to $10^{-100}$ and equally spaced on the log-log-scale (Table \ref{tab:thr}).
As we report the optimal threshold based on the test set, the AUC for ``PRS-max'' is an upper bound of the AUC for the C+T method.

T-Trees (\textit{Trees inside Trees}) is an algorithm derived from random forests \cite[]{breiman2001random} that takes into account the correlation structure among the genetic markers implied by linkage disequilibrium in GWAS data \cite[]{botta2014exploiting}. We use the same parameters as reported in Table 4 of \cite{botta2014exploiting}, except that we use 100 trees instead of 1000 because using 1000 trees provides a minimal increase of AUC while requiring a disproportionately long processing time (e.g.\ AUC of 81.5\% instead of 81\%, data not shown). %We call this method ``T-Trees'' in the results.

Finally, for the penalized logistic regression, we find regression coefficients $\beta_0$ and $\beta$ that minimize the following regularized loss function $$L(\lambda, \alpha) = \underbrace{ -\sum_{i=1}^n \left( y_i \log\left(p_i\right) + (1 - y_i) \log\left(1 - p_i\right) \right) }_\text{Loss function}   +   \underbrace{ \lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right) }_\text{Penalization} ~,$$ where $p_i=1/\left(1+\exp\left(-(\beta_0 + x_i^T\beta)\right)\right)$, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status to predict, $\lambda$ and $\alpha$ are two regularization hyper-parameters that need to be chosen.
Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (``ridge'', \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$ in the previous equation);
the L1-regularization (``lasso'', \cite{tibshirani1996regression}) forces some of the coefficients to be equal to zero and can be used as a means of variable selection, leading to sparse models (corresponds to $\alpha = 1$);
the L1- and L2-regularization (``elastic-net'', \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $m \gg n$ situation ($m$: number of SNPs), or any situation involving many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}. In this study, we use an embedded grid search over $\alpha \in \{1, 0.5, 0.05, 0.001\}$.

To fit this penalized logistic regression, we use an efficient algorithm \cite[]{friedman2010regularization,tibshirani2012strong,zeng2017efficient} from which we derived our own implementation in R package bigstatsr.
This type of algorithm builds predictions for many values of $\lambda$, which is called a ``regularization path''. To obtain an algorithm free of the choice of this hyper-parameter $\lambda$, we developed a procedure that we call Cross-Model Selection and Averaging (CMSA, figure \ref{fig:CMSA}).
Because of L1-regularization, the resulting vectors of coefficients are sparse and can be used to make a PRS based on a \textit{linear} combination of allele counts. We refer to this method as ``logit-simple'' in the results section.

@@

\subsection{Reproduciblity}

All the code used in this paper along with results such as figures and tables, are available as HTML R notebooks in the Supplementary Materials.
TODO: déplacer dans code availbility après ackn + ajouter les results + le tuto

@@

We compared penalized logistic regression (``logit-simple'') with the C+T method (``PRS'') using whole-genome simulations of scenario \textnumero1 (Table \ref{tab:simus}).

When simulating a model with 30 causal SNPs and an heritability of 80\%, penalized logistic regression provides AUC of 93\%, nearly reaching the maximum achievable AUC of 94\%, whereas AUC values obtained with C+T method range between 83\% and 90\% (Figures \ref{fig:main-AUC-logit} and \ref{fig:main-AUC-PRS}). 
Moreover, penalized logistic regression consistently provides higher predictive performance than the C+T method across all scenarios we considered, excepted in some cases of high polygenicity or small sample size where all methods perform poorly (AUC values below 60\% -- figures \ref{fig:main-AUC-ntrain} and \ref{fig:supp-AUC-logit}).

Method ``logit-simple'' provides particularly higher predictive performance than ``PRS-max'' when there are correlations between predictors, i.e.\ when we choose causal SNPs to be in the HLA region. In this situation, the mean AUC reaches 92.5\% with the ``logit-simple'' approach and 84\% with ``PRS-max'' (Figure \ref{fig:main-AUC-logit}).

Note that for the simulations we do not report results in terms of partial AUC because partial AUC values have a Spearman correlation of 98\% with the AUC results for all methods (Figure \ref{fig:supp-AUC-corr}).

@@

In practice, a particular value of the threshold of inclusion of SNPs should be chosen for the C+T method and this choice can dramatically impact the predictive performance of C+T. For example, in a model with only 30 causal SNPs, AUC ranges from less than 60\% when using all SNPs passing clumping to 90\% if choosing the optimal p-value threshold (Figures \ref{fig:main-AUC-PRS} and \ref{fig:supp-AUC-PRS}).

@@

Method ``logit-simple'' that automatically chooses hyper-parameter $\lambda$ provides similar predictive performance than the best predictive performance achieved ... of the implementation of R package biglasso \cite[]{zeng2017efficient}, only slightly better for biglasso, which is likely due to over-fitting when reporting the best prediction (Figure \ref{fig:supp-biglasso}).
TODO: juste dire qu'on est quasiment le max, sans mentionner biglasso. SIMILAR

@@

We tested the T-Trees method in scenario \textnumero1.
As compared to ``logit-simple'', T-Trees perform worse in terms of predictive ability, while taking much longer to run and making more complex predictive models because T-Trees use more predictors and non-linear effects (Figure \ref{fig:supp-ttrees}). 
Even when simulating a ``fancy'' model in which there are dominant and interaction-type effects that T-Trees should be able to handle, AUC is still lower when using T-Trees than when using ``logit-simple'' (Figure \ref{fig:supp-ttrees}).

We also compared the two penalized logistic regressions in scenario \textnumero1, ``logit-simple'' and ``logit-triple'' that uses additional features (variables) coding for recessive and dominant effects.
Predictive performance of ``logit-triple'' are nearly as good as ``logit-simple'' when there are only linear effects (differences of AUC are always smaller than 2\%) and can lead to significantly greater results when there are also dominant and interactions effects (Figures \ref{fig:supp-triple} and \ref{fig:supp-AUC-triple}). For the ``fancy model'', ``logit-triple'' provides AUC values at least 3.5\% higher than ``logit-simple'', excepted when there are 3000 causal SNPs.
Yet, the ``triple'' solution takes 2-3 times as much time to run and requires 3 times as much disk storage as the ``simple'' solution.

@@

First, when reproducing simulations of scenario \textnumero1 using chromosome 6 only (scenario \textnumero2), the predictive performance of ``logit-simple'' always increase (Figure \ref{fig:supp-AUC-chr6-all-r2}). 
There is particularly a large increase when simulating 3000 causal SNPs: AUC from the ``logit-simple'' increases from 60\% to nearly 80\% for Gaussian effects and a disease heritability of 80\%.
On the contrary, when simulating only 30 or 300 causal SNPs on the corresponding dataset, AUC of the ``PRS-max'' does not increase, and even decreases for an heritability of 80\% (Figure \ref{fig:supp-AUC-chr6-all-r2}). 
Secondly, when varying the training size (scenario \textnumero3), we report an increase of AUC when increasing the training size, with a faster increase of AUC provided by ``logit-simple'' as compared to ``PRS-max'' (Figure  \ref{fig:main-AUC-ntrain}).

@@

Joint logistic regressions also provide higher AUC values for the Celiac data: 88.7\% with ``logit-simple'' and 89.1\% with ``logit-triple'' as compared to 82.5\% with the C+T method.
The relative increase in partial AUC, for specificities larger than 90\%, is even larger (42\% and 47\%) with partial AUC values of 0.0411, 0.0426 and 0.0289 obtained with ``logit-simple'', ``logit-triple'' and the C+T method, respectively.
Moreover, logistic regressions use less predictors, respectively 1570, 2260 and 8360 (Table \ref{tab:results-celiac}, figures \ref{fig:celiac-roc} and supplementary notebook ``results-celiac'').
%Note that for the C+T method, we still report the best result among 102 p-value thresholds.
In terms of computation time, we show that the ``logit-simple'' method, while learning jointly on all SNPs at once and testing different hyper-parameter values, is almost as fast as the C+T method (190 vs 130 seconds), and the ``logit-triple'' takes less than twice as long as the ``logit-simple'' (296 vs 190 seconds).

\begin{table}[h]
\caption{Results for the real Celiac dataset. The results are averaged over 100 runs where the training step is randomly composed of 12,000 individuals. In the parentheses is reported the standard deviation of $10^5$ bootstrap samples of the mean of the corresponding variable. Results are reported with 3 significant digits.\label{tab:results-celiac}}
\vspace*{0.5em}
\centering
\begin{tabular}{|l|c|c|c|c|}
  \hline
Method & AUC & pAUC & \# predictors & Execution time (s) \\
  \hline
PRS-max & 0.825 (0.000664) & 0.0289 (0.000187) & 8360 (744) & 130 (0.143) \\ 
logit-simple & 0.887 (0.00061) & 0.0411 (0.000224) & 1570 (46.4) & 190 (1.21) \\ 
logit-triple & 0.891 (0.000628) & 0.0426 (0.000219) & 2260 (56.1) & 296 (2.03) \\ 
   \hline
\end{tabular}
\end{table}

@@

In this comparative study, we present a computationally efficient implementation of penalized logistic regression. 
This model can be used to build polygenic risk scores based on very large individual-level SNP datasets such as the UK biobank \cite[]{bycroft2017genome}. 
In agreement with previous work \cite[]{abraham2013performance}, we show that jointly estimating SNP effects has the potential to substantially improve predictive performance as compared to the standard C+T approach in which SNP effects are learned independently. 
Penalized logistic regression nearly always outperform the C+T method, and the benefits of using it are more pronounced with an increasing sample size or when causal SNPs are correlated with one another.

@@

The choice of hyper-parameter values is very important since it can greatly impact method performance. In the C+T method, there are two main hyper-parameters: the $r^2$ and the $p_T$ thresholds that control how stringent are the clumping and thresholding steps, respectively. 
The choice of the $r^2$ threshold of the clumping step is important. 
Indeed, on the one hand, choosing a low value for this threshold may discard informative SNPs that are correlated.
Yet, on the other hand, when choosing a high value for this threshold, too much redundant information would be included in the model, which would add some noise to the PRS.
Based on the simulations, we find that using a stringent threshold ($r^2 = 0.05$) leads to higher predictive performance, even when causal SNPs are correlated. It means that, in most cases, avoiding redundant information is more important than including all causal SNPs.
The choice of the $p_T$ threshold is also very important as it can greatly impact the predictive performance of the C+T method, which we confirm in this study \cite[]{ware2017heterogeneity}. 
In this paper, we reported the maximum AUC of 102 different p-value thresholds, a threshold that should normally be learned on the training set only. To our knowledge, there are no clear standard on how to choose these two critical hyper-parameters.

>>>>

\abstract{
Polygenic Risk Scores (PRS) consist in combining the information across many single-nucleotide polymorphisms (SNPs) in a score reflecting the genetic risk of developing a disease. PRS might have a major impact on public health, possibly allowing for screening campaigns to identify high-genetic risk individuals for a given disease.  
The ``Clumping+Thresholding'' (C+T) approach, which is the most common method to derive PRS, uses only univariate genome-wide association studies (GWAS) summary statistics, which makes it fast and easy to use. 
However, previous work showed that jointly estimating SNP effects for computing PRS has the potential to significantly improve the predictive performance of PRS as compared to C+T.

@@

In conclusion, our study demonstrates that penalized logistic regression can achieve more discriminative polygenic risk scores, while being applicable to large-scale individual-level data thanks to the implementation we provide in the R package bigstatsr.\\

\textbf{Contact:} \href{florian.prive@univ-grenoble-alpes.fr}{florian.prive@univ-grenoble-alpes.fr} \& \href{michael.blum@univ-grenoble-alpes.fr}{michael.blum@univ-grenoble-alpes.fr}
}

@@

\section{Introduction}

@@

Several methods have been developed to predict disease status, or more generally any phenotype, based on SNP information. A commonly used method often called ``P+T'' or ``C+T'' (which stands for ``Clumping and Thresholding'') is used to derive PRS from results of Genome-Wide Association Studies (GWAS) \cite[]{chatterjee2013projecting,dudbridge2013power,evans2009harnessing,purcell2009common,wray2007prediction}. This technique uses GWAS summary statistics only, allowing for a fast implementation of C+T. However, C+T also has several limitations; for instance, previous studies have shown that predictive performance of C+T is very sensitive to the threshold of inclusion of SNPs, depending on the disease architecture \cite[]{ware2017heterogeneity}.
Linear Mixed-Models (LMMs) are another widely-used method in fields such as plant and animal breeding or for predicting highly heritable quantitative human phenotypes such as height \cite[]{lello2017accurate,yang2010common}. Yet, models resulting from LMM, known e.g.\ as ``gBLUP'', are not optimal for predicting disease status based on genotypes \cite[]{abraham2013performance}. Moreover, these methods and their derivatives are often computationally very demanding, both in terms of memory and time required, which makes them unlikely to be used for prediction on very large datasets \cite[]{golan2014effective,maier2015joint,speed2014multiblup,zhou2013polygenic}.
Finally, statistical learning methods have also been used to derive PRS for complex human diseases by jointly estimating SNP effects. Such methods include joint logistic regression, Support Vector Machine (SVM) and random forests \cite[]{abraham2012sparsnp,abraham2014accurate,botta2014exploiting,okser2014regularized,wei2009disease}.

@@

To perform simulations, we use real genotype data and simulate new phenotypes. 
In order to make our comparison as comprehensive as possible, we compare different disease architectures by varying the number, size and location of causal effects as well as the disease heritability. We also compare two different models for simulating phenotypes, one with additive effects only, and one that combines additive, dominant and interaction-type effects.
Overall, we find that PLR consistently achieves higher predictive performance than the C+T and T-Trees methods while being as fast as C+T. This demonstrates the feasibility and relevance of this approach for PRS computation on large modern datasets.

@@

We use real genotypes of European individuals from a case-control study for celiac disease \cite[]{dubois2010multiple}. The composition of this dataset is presented in table \ref{tab:celiac-data}. Details of quality control and imputation for this dataset are available in \cite{prive2017efficient}. For simulations presented later, we first restrict this dataset to controls from UK in order to remove the genetic structure induced by the celiac disease status and population structure. This filtering process results in a sample of 7100 individuals (see supplementary notebook ``preprocessing''). 
We also use this dataset for real data application, in this case keeping all 15,155 individuals (4496 cases and 10,659 controls). Both datasets contain 281,122 SNPs.

@@

We simulate binary phenotypes using a Liability Threshold Model (LTM) with a prevalence of 30\% \cite[]{falconer1965inheritance}. We vary parameters of the simulations in order to match a range of genetic architecture from low to high polygenicity. 
This is achieved by varying the number of causal variants and their location (30, 300, or 3000 anywhere in all 22 autosomal chromosomes or 30 in the HLA region of chromosome 6), and the disease heritability (50\% or 80\%). 
%SNP effects are drawn from either a Gaussian or from a Laplace distribution.
Liability scores are computed either from an additive model with linear effects only (``ADD'') or a more complex model that combines additive, dominant and interaction-type effects (``COMP''). For model ``ADD'', we compute the liability score of the i-th individual $$y_i = \sum_{j\in S_\text{causal}} w_j \cdot \widetilde{G_{i,j}} ~~+~~ \epsilon_i~,$$ where $w_j$ are weights generated from a Gaussian or a Laplace distribution, $G_{i,j}$ is the allele count of individual $i$ for SNP $j$, $\widetilde{G_{i,j}}$ corresponds to its standardized version (zero mean and unit variance for all SNPs), $\epsilon$ follows a Gaussian distribution $N(0, 1 - h^2)$ and $S_\text{causal}$ is the set of causal SNPs.
For model ``COMP'', we simulate liability scores using additive, dominant and interaction-type effects (see Supplementary Materials).

@@

In this study, we compare three different types of methods: the C+T method, T-Trees and penalized logistic regression (PLR).

The C+T (Clumping + Thresholding) method directly derives a Polygenic Risk Score (PRS) from the results of Genome-Wide Associations Studies (GWAS). In GWAS, a coefficient of regression (i.e.\ the estimated effect size $\hat\beta_j$) is learned independently for each SNP $j$ along with a corresponding p-value $p_j$. The SNPs are first clumped (C) so that there remain only loci that are weakly correlated with one another (this set of SNPs is denoted $S_\text{clumping}$). Then, thresholding (T) consists in removing SNPs with p-values larger than a user-defined threshold $p_T$. Finally, the PRS for individual $i$ is defined as the sum of allele counts of the remaining SNPs weighted by the corresponding effect coefficients $$\rm{PRS}_i = \sum_{\substack{j \in S_\text{clumping} \\ p_j~<~p_T}} \hat\beta_j \cdot G_{i,j}~,$$ where $\hat\beta_j$ ($p_j$) are the effect sizes (p-values) learned from the GWAS. In this study, we mostly report scores for a clumping threshold at $r^2 > 0.2$ within regions of 500kb, but we also investigate thresholds of $0.05$ and $0.8$. We report three different scores of prediction: one including all the SNPs remaining after clumping (denoted ``C+T-all''), one including only SNPs remaining after clumping and that have a p-value under the GWAS threshold of significance ($p < 5 \cdot 10^{-8}$, ``C+T-stringent''), and one that maximizes the AUC (``C+T-max'') for 102 thresholds between $1$ and $10^{-100}$ (Table \ref{tab:thr}).
As we report the optimal threshold based on the test set, the AUC for ``C+T-max'' is an upper bound of the AUC for the C+T method.

T-Trees (\textit{Trees inside Trees}) is an algorithm derived from random forests \cite[]{breiman2001random} that takes into account the correlation structure among the genetic markers implied by linkage disequilibrium in GWAS data \cite[]{botta2014exploiting}. We use the same parameters as reported in Table 4 of \cite{botta2014exploiting}, except that we use 100 trees instead of 1000. Using 1000 trees provides a minimal increase of AUC while requiring a disproportionately long processing time (e.g.\ AUC of 81.5\% instead of 81\%, data not shown). %We call this method ``T-Trees'' in the results.

Finally, for the penalized logistic regression (PLR), we find regression coefficients $\beta_0$ and $\beta$ that minimize the following regularized loss function $$L(\lambda, \alpha) = \underbrace{ -\sum_{i=1}^n \left( y_i \log\left(p_i\right) + (1 - y_i) \log\left(1 - p_i\right) \right) }_\text{Loss function}   +   \underbrace{ \lambda \left((1-\alpha)\frac{1}{2}\|\beta\|_2^2 + \alpha \|\beta\|_1\right) }_\text{Penalization} ~,$$ where $p_i=1/\left(1+\exp\left(-(\beta_0 + x_i^T\beta)\right)\right)$, $x$ is denoting the genotypes and covariables (e.g.\ principal components), $y$ is the disease status to predict, $\lambda$ and $\alpha$ are two regularization hyper-parameters that need to be chosen.
Different regularizations can be used to prevent overfitting, among other benefits: the L2-regularization (``ridge'', \cite{hoerl1970ridge}) shrinks coefficients and is ideal if there are many predictors drawn from a Gaussian distribution (corresponds to $\alpha = 0$ in the previous equation);
the L1-regularization (``lasso'', \cite{tibshirani1996regression}) forces some of the coefficients to be equal to zero and can be used as a means of variable selection, leading to sparse models (corresponds to $\alpha = 1$);
the L1- and L2-regularization (``elastic-net'', \cite{zou2005regularization}) is a compromise between the two previous penalties and is particularly useful in the $p \gg n$ situation ($p$: number of SNPs), or any situation involving many correlated predictors (corresponds to $0 < \alpha < 1$) \cite[]{friedman2010regularization}. In this study, we use an embedded grid search over $\alpha \in \{1, 0.5, 0.05, 0.001\}$.

To fit this penalized logistic regression, we use an efficient algorithm \cite[]{friedman2010regularization,tibshirani2012strong,zeng2017efficient} from which we derived our own implementation in R package bigstatsr.
This type of algorithm builds predictions for many values of $\lambda$, which is called a ``regularization path''. To obtain an algorithm free of the choice of this hyper-parameter $\lambda$, we developed a procedure that we call Cross-Model Selection and Averaging (CMSA, figure \ref{fig:CMSA}).
Because of L1-regularization, the resulting vectors of coefficients are sparse and can be used to make a PRS based on a \emph{linear} combination of allele counts. We refer to this method as ``PLR'' in the results section.

@@

@@

We compared penalized logistic regression (PLR) with the C+T method using whole-genome simulations of scenario \textnumero1 (Table \ref{tab:simus}).

When simulating a model with 30 causal SNPs and an heritability of 80\%, PLR provides AUC of 93\%, nearly reaching the maximum achievable AUC of 94\%, whereas AUC values obtained with C+T range between 83\% and 90\% (Figures \ref{fig:main-AUC-logit} and \ref{fig:main-AUC-PRS}). 
Moreover, PLR consistently provides higher predictive performance than C+T across all scenarios we considered, excepted in some cases of high polygenicity or small sample size where all methods perform poorly (AUC values below 60\% -- figures \ref{fig:main-AUC-ntrain} and \ref{fig:supp-AUC-logit}).

PLR provides particularly higher predictive performance than C+T when there are correlations between predictors, i.e.\ when we choose causal SNPs to be in the HLA region. In this situation, the mean AUC reaches 92.5\% for PLR and 84\% for ``C+T-max'' (Figure \ref{fig:main-AUC-logit}).

Note that, for the simulations, we do not report results in terms of partial AUC because partial AUC values have a Spearman correlation of 98\% with the AUC results for all methods (Figure \ref{fig:supp-AUC-corr}).

@@

In practice, a particular value of the threshold of inclusion of SNPs should be chosen for the C+T method and this choice can dramatically impact the predictive performance of C+T. For example, in a model with only 30 causal SNPs, AUC ranges from less than 60\% when using all SNPs passing clumping to 90\% \emph{if} choosing the optimal p-value threshold (Figures \ref{fig:main-AUC-PRS} and \ref{fig:supp-AUC-PRS}).

@@

Our implementation of PLR that automatically chooses hyper-parameter $\lambda$ provides similar predictive performance than the best predictive performance of 100 models corresponding to different values of $\lambda$ (Figure \ref{fig:supp-biglasso}).

@@

We tested the T-Trees method in scenario \textnumero1.
As compared to PLR, T-Trees perform worse in terms of predictive ability, while taking much longer to run and making more complex predictive models because T-Trees use more predictors and non-linear effects (Figure \ref{fig:supp-ttrees}). 
Even when simulating with model ``COMP'' in which there are dominant and interaction-type effects that T-Trees should be able to handle, AUC is still lower when using T-Trees than when using PLR (Figure \ref{fig:supp-ttrees}).

We also compared the two penalized logistic regressions in scenario \textnumero1, PLR and PLR3 that uses additional features (variables) coding for recessive and dominant effects.
Predictive performance of PLR3 are nearly as good as PLR when there are only addtive effects (differences of AUC are always smaller than 2\%) and can lead to significantly greater results when there are also dominant and interactions effects (Figures \ref{fig:supp-triple} and \ref{fig:supp-AUC-triple}). For model ``COMP'', PLR3 provides AUC values at least 3.5\% higher than PLR, excepted when there are 3000 causal SNPs.
Yet, PLR3 takes 2-3 times as much time to run and requires 3 times as much disk storage as PLR.

@@

First, when reproducing simulations of scenario \textnumero1 using chromosome 6 only (scenario \textnumero2), the predictive performance of PLR always increase (Figure \ref{fig:supp-AUC-chr6-all-r2}). 
There is particularly a large increase when simulating 3000 causal SNPs: AUC from PLR increases from 60\% to nearly 80\% for Gaussian effects and a disease heritability of 80\%.
On the contrary, when simulating only 30 or 300 causal SNPs with the corresponding dataset, AUC of ``C+T-max'' does not increase, and even decreases for an heritability of 80\% (Figure \ref{fig:supp-AUC-chr6-all-r2}). 
Secondly, when varying the training size (scenario \textnumero3), we report an increase of AUC with a larger training size, with a faster increase of AUC provided with PLR as compared to ``C+T-max'' (Figure  \ref{fig:main-AUC-ntrain}).

@@

Joint logistic regressions also provide higher AUC values for the Celiac data: 88.7\% with PLR and 89.1\% with PLR3 as compared to 82.5\% with ``C+T-max''.
The relative increase in partial AUC, for specificities larger than 90\%, is even larger (42\% and 47\%) with partial AUC values of 0.0411, 0.0426 and 0.0289 obtained with PLR, PLR3 and ``C+T-max'', respectively.
Moreover, logistic regressions use less predictors, respectively 1570, 2260 and 8360 (Table \ref{tab:results-celiac}, figure \ref{fig:celiac-roc} and supplementary notebook ``results-celiac'').
In terms of computation time, we show that PLR, while learning jointly on all SNPs at once and testing four different hyper-parameter values, is almost as fast as the C+T method (190 vs 130 seconds), and PLR3 takes less than twice as long as PLR (296 vs 190 seconds).

\begin{table}[h]
\caption{Results for the real Celiac dataset. The results are averaged over 100 runs where the training step is randomly composed of 12,000 individuals. In the parentheses is reported the standard deviation of $10^5$ bootstrap samples of the mean of the corresponding variable. Results are reported with 3 significant digits.\label{tab:results-celiac}}
\vspace*{0.5em}
\centering
\begin{tabular}{|l|c|c|c|c|}
  \hline
Method & AUC & pAUC & \# predictors & Execution time (s) \\
  \hline
C+T-max & 0.825 (0.000664) & 0.0289 (0.000187) & 8360 (744) & 130 (0.143) \\ 
PLR & 0.887 (0.00061) & 0.0411 (0.000224) & 1570 (46.4) & 190 (1.21) \\ 
PLR3 & 0.891 (0.000628) & 0.0426 (0.000219) & 2260 (56.1) & 296 (2.03) \\ 
   \hline
\end{tabular}
\end{table}

@@

In this comparative study, we present a computationally efficient implementation of penalized logistic regression (PLR). 
This model can be used to build polygenic risk scores based on very large individual-level SNP datasets such as the UK biobank \cite[]{bycroft2017genome}. 
In agreement with previous work \cite[]{abraham2013performance}, we show that jointly estimating SNP effects has the potential to substantially improve predictive performance as compared to the standard C+T approach in which SNP effects are learned independently. 
PLR nearly always outperform the C+T method, and the benefits of using PLR are more pronounced with an increasing sample size or when causal SNPs are correlated with one another.

@@

The choice of hyper-parameter values is very important since it can greatly impact method performance. In the C+T method, there are two main hyper-parameters: the $r^2$ and the $p_T$ thresholds that control how stringent are the clumping and thresholding steps, respectively. 
The choice of the $r^2$ threshold of the clumping step is important. 
Indeed, on the one hand, choosing a low value for this threshold may discard informative SNPs that are correlated.
Yet, on the other hand, when choosing a high value for this threshold, too much redundant information would be included in the model, which would add some noise to the PRS.
Based on the simulations, we find that using a stringent threshold ($r^2 = 0.05$) leads to higher predictive performance, even when causal SNPs are correlated. It means that, in most cases, avoiding redundant information is more important than including all causal SNPs.
The choice of the $p_T$ threshold is also very important as it can greatly impact the predictive performance of the C+T method, which we confirm in this study \cite[]{ware2017heterogeneity}. 
In this paper, we reported the maximum AUC of 102 different p-value thresholds, a threshold that should normally be learned on the training set only. To our knowledge, there is no clear standard on how to choose these two critical hyper-parameters.

On the contrary, in the penalized logistic regression presented here, we developed an automatic procedure called Cross-Model Selection and Averaging (CMSA) that releases investigators from the burden of choosing hyper-parameter $\lambda$ that accounts for the amount of regularization used in the model. Not only this procedure provides near-optimal results, but it also accelerates the model training thanks to the development of an early stopping criterion. 
Usually, cross-validation is used to choose hyper-parameter values and then the model is trained again with these particular hyper-parameter values \cite[]{Hastie2008,wei2013large}. Yet, performing cross-validation and retraining the model is computationally demanding; CMSA offers a less burdensome alternative. 
Concerning hyper-parameter $\alpha$ that accounts for the relative importance of the L1 and L2 regularizations, we use a grid  search directly embedded in the CMSA procedure. 
%Our penalized logistic regression implementation does not allow for setting $\alpha = 0$, a model that would assume the infinitesimal model. Yet, if using e.g.\ $\alpha = 0.001$, it would lead to a similar model.
